---
title: "The biobtreeR users guide"
author: "Tamer Gür"
graphics: no
package: biobtreeR
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{The biobtreeR users guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Introduction

The biobtreeR package provides an interface to [biobtree](https://github.com/tamerh/biobtree) tool which covers and process efficently large set of bioinformatics datasets in local computer and provide search and mappings functionalities for genomic research.

# Getting started 

## Build data
biobtreeR process and store the datasets locally by downloading from their public locations. To build the datasets `bbBuildData` is used. And once this process is completed 
it does not require to perform again unless to work on new datasets or fetching the latest version of the datasets. Following examples shows the example usages, 


```{r}
library(biobtreeR)
```

```{r}
bbBuildData(datasets = "hgnc") 
```


```{r eval=FALSE}
# build default datasets taxonomy ensembl(homo_sapiens) uniprot(reviewed) hgnc go eco efo
# duration vary to network speed
bbBuildData() 

# build only specific datasets
bbBuildData(datasets="uniprot,hgnc")

# build both mouse and human genomes in ensembl insted of default human
bbBuildData(genome="homo_sapiens,mus_musculus")

# build genomes from ensembl genomes fungi
bbBuildData(datasets="+ensembl_fungi",genomes="saccharomyces_cerevisiae")

# bacteria genomes with pattern
bbBuildData(datasets="+ensembl_bacteria",genomePattern="serovar_infantis,serovar_virchow")

```


## Start/stop server

Once the building data completed biobtree server starts with `bbStart` which allows performing queries and provide web interface. Once the analyses has been done server can be stopped `bbStop` 

```{r}
bbStart()
```


## List available Ensembl genomes 

In order to use in the build process all the available Ensembl genomes names can be listed with with `bbListGenomes` function. 

**List ensembl genomes**

```{r}
head(bbListGenomes("ensembl"))
```

**List ensembl genomes bacteria**

```{r}
head(bbListGenomes("ensembl"))
```

**List ensembl genomes fungi**

```{r}
head(bbListGenomes("ensembl_fungi"))
```

**List ensembl genomes metazoa**

```{r}
head(bbListGenomes("ensembl_metazoa"))
```

**List ensembl genomes plants**

```{r}
head(bbListGenomes("ensembl_plants"))
```

**List ensembl genomes protists**

```{r}
head(bbListGenomes("ensembl_protists"))
```


## Search identifiers and keywords

Searching dataset identfiers and keywords such as gene name, species name and accession is performed with bbSearch function by passing comma seperated terms.

```{r}
bbSearch("cas,shh")
```

```{r eval = FALSE}
bbSearch("tpi1,vav_human,homo_sapiens,ENST00000297261,10090")
```

If source parameter is passed search performed within the dataset.

```{r}
bbSearch("cas,HGNC:12009","hgnc")
```

The bbURL function is used to get the unique URL of dataset identifier 

```{r}
bbURL("HGNC:12009","hgnc")
```

## Chain mapping and filtering 
Mapping and filtering queries are performed via bibobtree's mapping query syntax which allowed chain mapping and filtering capability from identifiers to target identifiers or attributes. Mapping query syntax consist of single or multiple mapping queries in the format `map(dataset_id).filter(Boolean query expression).map(...).filter(...)...` and allow performing chain mapping among datasets. For mapping queries bbMapFilter function is used, for instance in following example, maps protein to its go terms and in the second query mapping has been done with filter.

```{r eval = FALSE}
bbMapFilter("AT5G3_HUMAN",'map(go)',attrs = "type")
```

```{r eval = FALSE}
bbMapFilter("AT5G3_HUMAN",'map(go).filter(go.type=="biological_process")',attrs = "type")
```

In the example for the first parameter single protein accession has been used but similar with bbSearch functions multiple identifers or keywords can be used. In the last query type attribute was used to filter mapping only with biological process go terms. Dataset attributes are used in the filters starts with their dataset name as in the above example it starts with `go.`

In order use in filter expressions, each datasets attributes lists with `bbAttr`function via sample identifier. For instance following query shows gene ontology attributes.

```{r eval = FALSE}
bbAttr("GO:0002020",'go')
```

# Example Use cases

In this section mapping functionality will be discussed in detail with example use cases for gene and protein centric use cases.

## Gene centric use cases

Ensembl, Ensembl Genomes and HGNC datasets are used for gene related data. One of the most common gene related dataset identfiers are `ensembl`,`hgnc`,`transcript`,`exon`. Let's start with listing their attiributes,

```{r}
bbAttr("tpi1","hgnc")
```

```{r eval = FALSE}
bbAttr("tpi1","ensembl")
```

```{r eval = FALSE}
bbAttr("ENST00000297261","transcript")
```

```{r eval = FALSE}
bbAttr("ENSE00001146308","exon")
```

Note that there are several other gene related datasets without attributes and can be used in mapping queries such as probesets, genebank and entrez etc. Full dataset list can be discovered with `bbConfig`. Now lets build example mapping queries,

**Map gene names to Ensembl transcript identifiers**

```{r eval = FALSE}
res<-bbMapFilter("ATP5MC3,TP53",'map(transcript)')
head(res)
```

**Map gene names to exon identifiers and retrieve the region**

```{r eval = FALSE}
res<-bbMapFilter("ATP5MC3,TP53",'map(transcript).map(exon)',attrs = "seq_region_name")
head(res)
```


**Get all Ensembl human identifiers and gene names on chromosome Y with lncRNA type**
```{r eval = FALSE}
res<-bbMapFilter("homo_sapiens",'map(ensembl).filter(ensembl.seq_region_name=="Y" && ensembl.biotype=="lncRNA")',attrs = "name")
head(res)
```


**Get all Ensembl human identifiers and gene names within or overlapping range**
```{r eval = FALSE}
bbMapFilter("9606",'map(ensembl).filter((114129278>ensembl.start && 114129278<ensembl.end) || (114129328>ensembl.start && 114129328<ensembl.end))',attrs = "name")
```

In the above example as a first parameter taxonomy identifier is used instead of specifying as homo sapiens like in the previous example. Both of these usage are equivalent and produce same output as homo sapiens refer to taxonomy identifer 9606.

**Map Affymetrix identifiers to Ensembl identifiers and gene names**

```{r eval = FALSE}
bbMapFilter("202763_at,213596_at,209310_s_at",source ="affy_hg_u133_plus_2" ,'map(transcript).map(ensembl)',attrs = "name")
```

Note that all mappings can be done with opposite way, for instance from gene name to Affymetrix identifiers mapping is performed following way

```{r eval = FALSE}
bbMapFilter("CASP3,CASP4",'map(transcript).map(affy_hg_u133_plus_2)')
```

**Retrieve all the human gene names which contains TTY**
```{r eval = FALSE}
res<-bbMapFilter("homo sapiens",'map(ensembl).filter(ensembl.name.contains("TTY"))',attrs = "name")
head(res)
```

**Map gene to its ortholog identifiers**

```{r eval = FALSE}
res<-bbMapFilter("shh",'map(ortholog)')
head(res)
```

**Map gene to its paralog**

```{r eval = FALSE}
bbMapFilter("fry",'map(paralog)',attrs = "name",showInputColumn = TRUE)
```


**Map ensembl identifier or gene name to the entrez identifier**
```{r eval = FALSE}
bbMapFilter("ENSG00000139618,shh" ,'map(entrez)')
```

**Map refseq identifiers to hgnc identifiers**

```{r}
bbMapFilter("NM_005359,NM_000546",'map(hgnc)',attrs = "symbols")
```


## Protein centric use cases

Uniprot is used for protein related dataset such as protein identifiers, accession, sequence, features, variants, and mapping information to other datasets. Let's list some protein related datasets attributes and then execute example queries similary with gene centric examples,

```{r eval = FALSE}
bbAttr("p15498","uniprot")
```

```{r eval = FALSE}
bbAttr("p15498_f12","ufeature")
```

```{r eval = FALSE}
bbAttr("1gzh","pdb")
```

```{r eval = FALSE}
bbAttr("IPR017907","interpro")
```


**Map gene names to reviewed uniprot identifiers**

```{r eval = FALSE}
bbMapFilter("msh6,stk11,bmpr1a,smad4,brca2","map(uniprot).filter(uniprot.reviewed)",source ="ensembl")
```

**Filter proteins by sequence mass and retrieve protein sequences**

```{r eval = FALSE}
bbMapFilter("clock_human,shh_human,aicda_human,at5g3_human,p53_human","filter(uniprot.sequence.mass > 45000)" ,attrs = "sequence$mass,sequence$seq")
```

**Helix type feature locations of a protein**

```{r eval = FALSE}
bbMapFilter("shh_human",'map(ufeature).filter(ufeature.type=="helix")' ,attrs = "location$begin,location$end")
```

**All protein identifiers and sequences of childs of Salmonella species**

```{r eval = FALSE}
res<-bbMapFilter("Salmonella enterica subsp. enterica","map(taxchild).map(uniprot)",attrs = "sequence$seq")
head(res)
```

**Get all variation identifiers from a gene with given condition**

```{r eval = FALSE}
bbMapFilter("tp53",'map(uniprot).map(ufeature).filter(ufeature.original=="I" && ufeature.variation=="S").map(variantid)',source = "hgnc")
```

# Clear data

```{r}
bbClearData()
```


# Session Info

```{r}
sessionInfo()
```
